Making Higher-Order Superposition Work

### 2022-03-25 ehoh2/ehoh2.tex
eval

>>>>

  \section{Preprocessing, Calculus, and Extensions}
  \label{sec:ehoh2:calculus}

  Ehoh's simple $\lambda$-free higher-order calculus performed well on
  Sledgehammer problems and formed a promising stepping stone to full higher-order
  logic (Sect.~\ref{sec:ehoh:evaluation}). When implementing support for full higher-order
  logic, we were guided by efficiency and gracefulness with respect to Ehoh's calculus rather
  than completeness. Whereas Zipperposition provides both complete and incomplete
  modes, \ehohii{} only offers incomplete modes.

  \ourpara{Preprocessing} Our experience with Zipperposition showed the importance
  of flexibility in preprocessing the higher-order problems
  (Sect.~\ref{sec:ho-tech:preprocessing}). Therefore, we implemented a flexible
  preprocessing module in \ehohii{}.

  To maintain compatibility with Ehoh, \ehohii{} can optionally transform all
  $\lambda$-abstractions into named functions. This process is called
  \emph{$\lambda$-lifting} \cite{rjmh-82-lifting}. \ehohii{} also removes all
  occurrences of Boolean subterms (other than $\itrue, \ifalse$, and free variables)
  in higher-order contexts using a FOOL-like transformation \cite{kotelnikov-16-fool}.
  For example, the formula $\cst{f}(\cst{p} \iand \cst{q}) \ieq \cst{a}$ becomes
  $(\cst{p} \iand \cst{q} \iimplies \cst{f}(\itrue) \ieq \cst{a}) \iand
  (\inot(\cst{p} \iand \cst{q}) \iimplies \cst{f}(\ifalse) \ieq \cst{a})$.

  \looseness=-1
  Many TPTP problems use the \texttt{definition} role to denote the definitions of
  symbols. \ehohii{} can treat the definition axioms as rewrite rules, and replace
  all occurrences of defined symbols during preprocessing. Furthermore, during
  SInE \cite{hv-2011-sine} axiom selection, it can always
  include the defined symbol in the trigger relation.

  \ourpara{Calculus} \ehohii{} implements the same superposition
  calculus as Ehoh with three important changes. First, wherever Ehoh requires
  the most general unifier of terms, \ehohii{} enumerates unifiers from a finite subset of the CSU, as explained in
  Sect.~\ref{sec:ehoh2:unif-match-index}. Second, \ehohii{} uses versions of the KBO and
  LPO orders designed for $\lambda$-terms.
  % as described by Bentkamp et al.\ \cite{bbtv-21-full-ho-sup}.

  %TODO: @JASMIN: How do we cite the STTT article?
  The third difference is more subtle. One of the main features of Ehoh is
  \emph{prefix optimization} (Sect.~\ref{sec:ehoh:introduction}): A method which, given a demodulator $s \eq t$, allows
  to replace both applied and unapplied occurrences of $s$ by $t$ 
  by traversing only first-order subterms of a rewritable term. In $\lambda$-free setting this
  optimization is useful, but in presence of $\beta\eta$-normalization, shapes of
  terms can change drastically, making it much harder to
  track prefixes of terms. This is why we disabled the prefix optimization in \ehohii{}.
  To counterbalance this removal, we introduced
  the argument congruence rule \infname{AC} %(also called \infname{ArgCong})
  in \ehohii{}
  and enabled positive and negative functional extensionality (\infname{PE} and
  \infname{NE}) by default:

  \vskip\abovedisplayskip

  \noindent\hbox{}\hfill
    \namedinference{AC}{s \eq t \llor C}{s \, X \eq t \, X \llor C}%
  \hfill
    \namedinference{NE}{s \noteq t \llor C}{s \; (\cst{sk} \; \overline{X}) \noteq t \; (\cst{sk} \; \overline{X}) \llor C}%
  \hfill
    \namedinference{PE}{s \; X \eq t \; X \llor C}{s \eq t \llor C}%
  \hfill\hbox{}

  \vskip\belowdisplayskip

  \noindent
  \infname{AC} and \infname{NE} assume that $s$ and $t$ are of function
  type. In \infname{NE}, $\overline{X}$ denote all the free variables occurring in $s$ and
  $t$, and $\cst{sk}$ is a fresh Skolem symbol of the appropriate type. \infname{PE}
  has a side condition that $X$ may not appear in $s$, $t$, or $C$.

  \ourpara{Saturation} E's saturation procedure is based on the assumption that
  each attempt to perform an inference will either result in a single clause or
  fail due to one of the inference side conditions. Unification procedures that
  produce multiple substitutions break this invariant, and the saturation
  procedure needed to be adjusted.

  In Sect.~\ref{sec:ho-tech:infinite-branching}, we described a variant of the saturation
  procedure that elegantly supports interleaving computation of unifiers and
  scheduling
  inferences to be performed. Since completeness was
  not one of the \ehohii{} design goals, we did not implement this version of the
  saturation procedure. Instead, in places where previously a single unifier was
  expected, \ehohii{} consumes all elements of the iterator used for enumerating a
  unifier, converting them to clauses.

  \ourpara{Reasoning about Formulas} Even though most of the Boolean structure is
  removed during preprocessing, formulas can reappear at the top level of
  clauses during saturation. For example, after replacing $X$ with $\lambda x.
  \, \lambda y.\, x \iand y$, the clause $X \, \cst{p} \, \cst{q} \llor \cst{a}
  \eq \cst{b}$ becomes $(\cst{p} \iand \cst{q}) \llor \cst{a} \eq \cst{b}$. \ehohii{}
  converts every clause of the form $\varphi \llor C$, where $\varphi$ has a logic symbol
  as its head (or it is a (dis)equation between two formulas different than $\itrue$),
  to an explicitly quantified formula. Then, the clausification algorithm
  is invoked on the formula to restore the clausal structure. Zipperposition features
  more dynamic clausification modes, but for simplicity we decided
  not to implement them in \ehohii{}.

  The \osup{} calculus
  \cite{bbtv-21-full-ho-sup} includes many rules that act on Boolean subterms,
  which are necessary for completeness. Other than Boolean simplification rules,
  which use simple tautologies such as $\cst{p} \iand \itrue \iequiv\cst{p}$ to simplify terms, we
  have implemented none of the Boolean rules of this calculus in \ehohii{}. First, we have
  observed that complicated rules such as \infname{FluidBoolHoist} and
  \infname{FluidLoobHoist} are hardly ever useful in practice and usually only
  contribute to an uncontrolled increase in the proof state size. Second, simpler
  rules such as \infname{BoolHoist} can usually be simulated by pragmatic rules
  that perform Boolean and functional extensionality reasoning, described below.

  To make up for excluding Boolean rules, we use an incomplete, but more
  easily controllable and intuitive rule, called \emph{primitive instantiation}. This
  rule instantiates free predicate variables with approximations of formulas
  that are ground instances of this variable. We described this rule in Sect.~\ref{sect:bool:native}.
  In \ehohii{} it is implemented in a similar manner.

  \ehohii{}'s handling of the Hilbert choice operator is inspired by Leo-III's
  \cite{sb-21-leo3}. \ehohii{}
  recognizes clauses of the form $\neg P \, X \lor P \, (\cst{f} \, P)$ which
  essentially denote that $\cst{f}$ is a choice symbol. Then, when subterm
  $\cst{f} \, s$ is found during saturation, $s$ is used to instantiate the choice
  axiom for $\cst{f}$. 
  %
  Similarly, Leibniz equality is eliminated by
  recognizing clauses of the form  $\neg P \, \cst{a} \llor P \, \cst{b} \llor C$. These clauses are then
  instantiated with $P \mapsto \lambda x. \, x \eq \cst{a}$ and $P \mapsto \lambda
  x. \, x\noteq\cst{b}$, which results in $\cst{a} \eq \cst{b} \lor C$. Both rules
  are described in Sect.~\ref{sect:bool:native} in more detail.

  Finally, \ehohii{} treats induction axioms specially. Just like
  Zipperposition (Sect.~\ref{sec:ho-tech:formulas}), it abstracts literals
  from the goal clauses and uses these abstractions to instantiate induction
  axioms. Since Zipperposition supports dynamic calculus-level clausification,
  instantiation of induction axioms happens during saturation, when these axioms
  are processed. In \ehohii{}, this instantiation is performed statically,
  immediately after clausification. After \ehohii{} collects all the abstractions,
  it traverses the clauses and instantiates those that have applied variable of
  the same type as the abstraction.

  \ourpara{Extensionality} \ehohii{} takes a pragmatic approach to reason about
  functional and Boolean extensionality: It uses \emph{abstracting} rules (\infname{Abs} rules of Sect.~\ref{sect:bool:native}) 
  which simulate basic superposition calculus rules,
  but do not require unifiability of the partner terms in the inference. More
  precisely, assume a core inference needs to be performed between two
  $\beta$-reduced terms $u$ and $v$, such that they can be represented as
  $u=C[s_1, \ldots, s_n]$ and $v=C[t_1, \ldots, t_n]$, where $C$ is the most general
  common context (more precisely, green common context that does not go into the structure of $\lambda$-abstraction or applied variables \cite{bbtvw-21-sup-lam}) of $u$ and $v$, not all of
  $s_i$ and $t_j$ are free variables, and for at least one $i$, $s_i \not= t_i$,
  $s_i$ and $t_i$ are not (applied) free variables, and they are of Boolean or
  function type. Then, the conclusion is formed by taking the conclusion $D$ of
  the core inference rule (which would be created if $s$ and $t$ are unifiable)
  and adding literals $s_1 \noteq t_1 \llor \cdots \llor s_n \noteq t_n$.

  These rules are particularly useful because \ehohii{} has no rules that dynamically
  process Booleans in FOOL-like fashion, such as \infname{BoolHoist}. For example,
  given the clauses $\cst{f} \, (\cst{p} \iand \cst{q}) \eq \cst{a}$ and
  $\cst{g} \, (\cst{f} \, \cst{p}) \noteq \cst{b}$, the abstracting version of
  superposition rule would result in $\cst{g} \, \cst{a} \noteq \cst{b} \llor (\cst{p}
  \iand \cst{q}) \noteq \cst{p}$. In this way, the Boolean structure bubbles up to
  the top level and is further processed by clausification. We noticed that this
  alleviates the need for the other Boolean rules in practice.

  \section{Evaluation}
  \label{sec:ehoh2:eval}

  In this section, we try to answer two questions about \ehohii{}: \emph{How does
  \ehohii{} compare against other higher-order provers (including Ehoh)?} \emph{Does
  \ehohii{} introduce any overhead compared with Ehoh?} To answer
  these questions, we ran provers on problems from the TPTP library \cite{gs-17-tptp} and on
  benchmarks generated by Sledgehammer (SH) \cite{pb-12-sh}.
  The experiments were carried out on StarExec Miami
  \cite{sst-14-starexec} nodes equipped with Intel Xeon E5-2620 v4 CPU clocked at
  2.10~GHz. For the TPTP part, we used the CASC-28\footnote{\url{http://www.tptp.org/CASC/28/}} time limits:\ 120~s
  wallclock and 960~s CPU. For SH benchmarks and to answer the other question, we used Sledgehammer's
  default time limit:\ 30~s wallclock and CPU.

  \ourpara{Comparison with Other Provers}
  To answer the first question, we let
  \ehohii{} compete with the top contenders in the higher-order division of
  CASC-28:\ cvc5 \NumberOK{0.0.7},\footnote{\url{https://cvc5.github.io/}}
  Ehoh 2.7, Leo-III \NumberOK{1.6.6} \cite{sb-21-leo3}, Vampire 4.6
  \cite{br-20-full-sup-w-combs}, and Zipperposition 2.1 (Chapter~\ref{ch:ho-techniques}). 
  We also include Satallax 3.5
  \cite{cb-12-satallax}. We used all \NumberOK{2899} higher-order
  theorems in TPTP 7.5.0 as well as 5000 SH higher-order benchmarks
  originating from the Seventeen benchmark suite \cite{desharnais-et-al-202x}.
  On SH benchmarks,
  cvc5, Ehoh, \ehohii{}, Vampire, and Zipperposition were run using custom schedules provided by
  their developers, optimized for single-core usage and low timeouts.
  Otherwise, we used the corresponding CASC configurations.
  Note that Ehoh 2.7 is an updated version of Ehoh described in Chapter
  \ref{ch:ehoh}, which can parse not only $\lambda$-free but also full
  higher-order. It is only a syntactic extension as $\lambda$-abstractions are supported using $\lambda$-lifting \cite{rjmh-82-lifting}.
  We included two versions of Zipperposition:\ \emph{coop} uses Ehoh 2.7 as a backend to finish proof attempts,
  whereas \emph{uncoop} does not use this feature. The results
  are shown in Figure \ref{fig:all-provers}.

  Both Ehoh and \ehohii{} were run in the automatic scheduling mode. Compared to Ehoh,
  \ehohii{} features a redesigned module for automatic scheduling, it
  can use multiple CPU cores, and its heuristics have been trained better on higher-order problems.

  \ehohii{} dramatically improves the higher-order reasoning capabilities
  compared to Ehoh. It solves \NumberOK{20\%} more problems on TPTP benchmarks
  and \NumberOK{6\%} more problems on SH benchmarks, where Ehoh was already very
  successful.

  \ehohii{} was mainly designed as efficient backed to proof assistants. As such it 
  excels on SH benchmarks, outperforming the competition. On TPTP, it outperforms all higher-order provers other than Zipperposition-coop.
  If Zipperposition's Ehoh backend is disabled,
  \ehohii{} outperforms Zipperposition by a wide margin.
  This comparison is arguably fairer; after all,
  \ehohii{} does not use an older version of Zipperposition
  as a backend.
  %
  These results suggest that \ehohii{} already implements most of the necessary features
  for a high-performance higher-order prover
  but could benefit from the kind of fine-tuning that
  Zipperposition underwent in the last two years.

  Remarkably, there is a substantial disparity in the set of problems solved by
  \ehohii{} and Zipperposition-coop. The raw data show thats \ehohii{} solves
  \NumberOK{166} SH problems and \NumberOK{21} TPTP problems that Zipperposition-coop
  does not. The lower number of uniquely solved TPTP problems is most likely 
  due to Zipperposition's being heavily optimized on TPTP.

  \ourpara{Comparison with the First-Order E} Both Ehoh and \ehohii{} can be compiled
  in a mode that disables most of the higher-order reasoning. This mode is
  designed for users that are interested only in E's first-order capabilities and
  care a lot about performance.
  To answer the second
  evaluation question, about assessing overhead of \ehohii{},
  we chose all the \NumberOK{1138} unique problems used at CASC
  from 2019 to 2021 in first-order theorem division and ran Ehoh and \ehohii{}
  both in this first-order (FO) mode and in higher-order (HO) mode.

  We fixed a single
  configuration of options, because Ehoh and \ehohii{}'s
  automatic scheduling methods could select different configurations
  and we would not be measuring the overhead but the quality of the chosen
  configurations. We chose the \emph{boa} configuration (Sect.~\ref{sec:ehoh:heuristics}),
  which is the configuration that E~2.2 used the most often in its automatic scheduling mode.
  The results are shown in Figure~\ref{fig:ehoh}.

  Counterintuitively, the higher-order versions of both provers outperform
  the first-order counterparts. However, the differerence is so small that it can
  be attributed to the changes to memory layout that affect the order in which
  clauses are chosen. Similar effects are seen when comparing the first-order versions.
  We would expected the less heavily modified version of E, Ehoh, to perform better,
  but due to subtle effects \ehohii{} wins.

  \begin{figure}[t]
    \noindent\hbox{}\hfill
    \begin{minipage}[ht!]{.66\linewidth}
      \centering
      \def\arraystretch{1.1}%
      \relax{\begin{tabular}{l@{\hskip 1em}@{\hskip 0.5em}c@{\hskip 1em}c} \toprule

                              & TPTP       & SH        \\
        \midrule
        cvc5                  & 1931       & 2577      \\
        Ehoh                  & 2105       & 2611      \\
        \ehohii{}             & 2523       & {\bf 2766}\\
        Leo-III               & 2282       & 1601      \\
        Satallax              & 2320       & 1719      \\
        Vampire               & 2203       & 2240      \\
        Zipperposition-coop   & {\bf 2583} & 2754      \\
        Zipperposition-uncoop & 2483       & 2181      \\
        \bottomrule
      \end{tabular}}
      \caption{Comparison of higher-order provers}
      \label{fig:all-provers}
    \end{minipage}\hfill\hfill
    \begin{minipage}[ht!]{.30\linewidth}
      \centering
      \def\arraystretch{1.1}%
      \relax{\begin{tabular}{l@{\hskip 1em}@{\hskip 0.5em}c} \toprule
                        & TPTP     \\
        \midrule
        Ehoh -- FO      & 535      \\
        Ehoh -- HO      & 538      \\
        \ehohii{} -- FO & {\bf 539}\\
        \ehohii{} -- HO & {\bf 539}\\
        \bottomrule
      \end{tabular}}
      \caption{Evaluation of \ehohii{}'s overhead}
      \label{fig:ehoh}
    \end{minipage}\hfill\hfill
  \end{figure}

### 2022-03-25 scratch/paper.tex
unif
eval

\section{Unification, Matching, and Term Indexing}
\label{sec:ehoh2:unif-match-index}
% \begin{itemize}
%   \item Procedure that implements a variant of the algo described in Efficient
%   HO Unif paper.
%   \item Two oracles -- fixpoint and pattern unif (solid not implemented as
%   complexity/gain ratio is quite negative)
%   \item It is implemented iteratively, with subtle optimizations
%   \item Its iterator is completely contained -- iterators can be stored and
%   later waked up similarly to what we do in Zipperposition.
%   \item Mention that so far the prover fares extremely well with complicated
%   unification problems and that we do not consider implementing more complex
%   given clause algo in Ehoh.
%   \item Matching works only with algorithms that return a single matcher:
%   pattern matching and LFHOL matching : main optimization from E is preserved.
%   PDT algorithm also works with both kinds of terms LFHOL and pattern. Check if
%   something is a pattern is efficient!
%   \item Fingerprints and feature vectors as in Zipperposition.
%   \item To avoid huge refactorings iterator is used even if the problem to be
%   solved is first-order.  
% \end{itemize}

@@

\section{Preprocessing, Calculus, and Extensions}

\label{sec:ehoh2:calculus}
% \begin{itemize}  
  % \item The main goal is simplicity, efficiency and maximal compatibility with
  %     FO superposition.
  % \item More or less the same as LFHOL sup, but now computes CSUs instead
  %     of LFHOL MGUs (and (potentially) uses new orders).
  % \item Out of Boolean rules we have Boolean simplification and black-box clausification.
  %     Most importantly -- we have no Boolean hoisting rules : everything is dealt with
  %     ext-sup, and FOOL clausification! It simply works as good as Boolean hoisting.
  % \item IMPORTANT: Prefix optimization is disabled, now we rely on ArgCong and NegExt.
  % \item Many extensions:  ExtSup (With the same indexing), leibniz elimination, predicate enumeration,
  %     FlexElim, Choice instantiation, recognition of induction.
  % \end{itemize}

@@

These rules are particularly useful because \ehohii{} has no rules that dynamically
process Booleans in FOOL-like fashion, such as \infname{BoolHoist}. For example,
given the clauses $\cst{f} \, (\cst{p} \iand \cst{q}) \eq \cst{a}$ and
$\cst{g} \, (\cst{f} \, \cst{p}) \noteq \cst{b}$, the abstracting version of
superposition rule would result in $\cst{g} \, \cst{a} \noteq \cst{b} \llor (\cst{p}
\iand \cst{q}) \noteq \cst{p}$. In this way, the Boolean structure bubbles up to
the top level and is further processed by clausification. We noticed that this
aleviates the need for the other Boolean rules in practice.

@@

Remarkably, there is a substantial disparity in the set of problems solved by
\ehohii{} and Zipperposition-coop. The raw data show thats \ehohii{} solves
\NumberNOK{166} SH problems and \NumberNOK{21} TPTP problems that Zipperposition-coop
does not. The lower number of uniquely solved TPTP problems is most likely 
due to Zipperposition's being heavily optimized on TPTP.

\ourpara{Comparison with the First-Order E} Both Ehoh and \ehohii{} can be compiled
in a mode that disables most of the higher-order reasoning. This mode is
designed for users that are interested only in E's first-order capabilities and
care a lot about performance.
To answer the second
evaluation question, about assessing overhead of \ehohii{},
we chose all the \NumberNOK{1138} unique problems used at CASC
from 2019 to 2021 in first-order theorem division and ran Ehoh and \ehohii{}
both in this first-order (FO) mode and in higher-order (HO) mode.

>>>>

\section{Unification, Matching, and Term Indexing}
\label{sec:ehoh2:unif-match-index}

@@

\section{Preprocessing, Calculus, and Extensions}
\label{sec:ehoh2:calculus}

@@

These rules are particularly useful because \ehohii{} has no rules that dynamically
process Booleans in FOOL-like fashion, such as \infname{BoolHoist}. For example,
given the clauses $\cst{f} \, (\cst{p} \iand \cst{q}) \eq \cst{a}$ and
$\cst{g} \, (\cst{f} \, \cst{p}) \noteq \cst{b}$, the abstracting version of
superposition rule would result in $\cst{g} \, \cst{a} \noteq \cst{b} \llor (\cst{p}
\iand \cst{q}) \noteq \cst{p}$. In this way, the Boolean structure bubbles up to
the top level and is further processed by clausification. We noticed that this
alleviates the need for the other Boolean rules in practice.

@@

Remarkably, there is a substantial disparity in the set of problems solved by
\ehohii{} and Zipperposition-coop. The raw data show thats \ehohii{} solves
\NumberOK{166} SH problems and \NumberOK{21} TPTP problems that Zipperposition-coop
does not. The lower number of uniquely solved TPTP problems is most likely 
due to Zipperposition's being heavily optimized on TPTP.

\ourpara{Comparison with the First-Order E} Both Ehoh and \ehohii{} can be compiled
in a mode that disables most of the higher-order reasoning. This mode is
designed for users that are interested only in E's first-order capabilities and
care a lot about performance.
To answer the second
evaluation question, about assessing overhead of \ehohii{},
we chose all the \NumberOK{1138} unique problems used at CASC
from 2019 to 2021 in first-order theorem division and ran Ehoh and \ehohii{}
both in this first-order (FO) mode and in higher-order (HO) mode.
